\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-24 NAR_DRAFT.tex
remove cruft

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \citep{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\citep{Rubanova2018a}, although those differences are relatively small.
A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increasing sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

%\batcave resembles an empirical Bayes approach, but is not really empirical Bayes.
%In fact, variants are best described as sequential samples from a data generating distribution represented by the mutation profile.
%Early mutations are classified with low error, while later mutations are classified with less certainty.
%We show here that sequential updating of the dirichlet-multinomial posterior converges to the true data generating distribution using the set of early mutations.
%Along with the assumption that tumor evolution is approximately neutral, this allows us to generate a tumor and site-specific prior that sharpens the classification and calibration of later mutations.
%Thus, \batcave is best described as an algorithm trained on early mutations.

>>>>

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \citep{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\citep{Rubanova2018a}, although those differences are relatively small.
A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increase sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.
